The Galactic Midplane Is Not a Plane: Implications for Dynamical Analysis with {\em Gaia} Data and Beyond

### 2019-04-26 paper/ms.tex
cosmetic changes to derivation in section 2 - see comments in slack for other needed checks

While the formula for $J_{\phi}$ is phase independent, $J_R$ and $J_z$ are
both phase dependent. We consider two cases. First, at the point of maximum
amplitude or when $(v_R, v_z) = (0, 0)$ and second, at the point of maximum
velocity $(R, z) = (R_g, 0)$. For $J_R$ we have that,
\beq\label{eq:JR_epi_final}
    \begin{aligned}
        J_R &= \kappa A_R^2 && \text{if $v_R = 0$} \\
    J_R &= \frac{v_{R,\text{max}}^2}{2\kappa^2} && \text{if $R=R_g$} \text{.}
    \end{aligned}
\eeq
For $J_z$, we have that,
\beq\label{eq:Jz_epi_final}
    \begin{aligned}
        J_z &= \kappa A_z^2 && \text{if $v_z = 0$} \\
        J_z &= \frac{v_{z,\text{max}}^2}{2\kappa^2} && \text{if $z=0$} \text{.}
    \end{aligned}
\eeq

Now assume that the coordinates $(R, z, v_{\phi}, v_R, v_z)$ are offset by
$(\Delta R, \Delta z, \Delta v_{\phi}, \Delta v_R, \Delta v_z)$. Then, the
induced error in $J_{\phi}$ is,
\beq\label{eq:induced_Jphi}
\frac{\Delta J_{\phi}}{J_{\phi}} = \frac{\Delta R}{R_g}\text{.}
\eeq
For $J_R$, the induced error is,
\beq\label{eq:induced_JR}
\begin{aligned}
    \frac{\Delta J_R}{J_R} &= \frac{2\Delta R}{A_R} + 
        \left(\frac{\Delta v_R}{v_{R,\text{max}}}\right)^2 && \text{if $v_R = 0$} \\
    \frac{\Delta J_R}{J_R} &= \frac{2\Delta v_R}{v_{R,\text{max}}} + 
        \left(\frac{\Delta R}{A_R}\right)^2  && \text{if $R=R_g$} \text{.}
\end{aligned}
\eeq
For $J_{z}$, the induced error is,
\beq\label{eq:induced_Jz}
\begin{aligned}
    \frac{\Delta J_z}{J_z} &= \frac{2\Delta z}{A_z} + 
        \left(\frac{\Delta v_z}{v_{z,\text{max}}}\right)^2 && \text{if $v_z = 0$} \\
    \frac{\Delta J_z}{J_z} &= \frac{2\Delta v_z}{v_{z,\text{max}}} + 
        \left(\frac{\Delta z}{A_z}\right)^2  && \text{if $z=0$} \text{.}
\end{aligned}
\eeq

Since most of the time stars will be at maximum amplitude in both the $R$ and
$z$ components, we can assume for most stars that
\beq\label{eq:Ji_err_mosttime}
\begin{split}
\frac{\Delta J_{\phi}}{J_{\phi}} &= \frac{\Delta R}{R_g} \\
\frac{\Delta J_{R}}{J_{R}} &= \frac{2\Delta R}{A_R} \\
\frac{\Delta J_{z}}{J_{z}} &= \frac{2\Delta z}{A_z} \text{,}
\end{split}
\eeq
where we have also ignored second order terms.

In the remainder of this section, we compare our analytic estimates of the effect on actions of a midplane offset against numerical calculations. Detailed worked on the effect of velocity offsets on actions is deferred to future work, as we discuss in Section~\ref{ssec:lsr_var}.

>>>>

While the formula for $J_{\phi}$ is phase independent, $J_R$ and $J_z$ are
both phase dependent via Eq. \eqref{eq:orbits_epi}. We consider two cases. First, at the point of maximum
amplitude or when $(v_R, v_z) = (0, 0)$ and second, at the point of maximum
velocity $(R, z) = (R_g, 0)$. For $J_R$ we have that,
\beq\label{eq:JR_epi_final}
    \begin{aligned}
        J_R &= \kappa A_R^2 && \text{if $v_R = 0$} \\
    J_R &= \frac{v_{R,\text{max}}^2}{2\kappa^2} && \text{if $R=R_g$} \text{.}
    \end{aligned}
\eeq
For $J_z$, we have that,
\beq\label{eq:Jz_epi_final}
    \begin{aligned}
        J_z &= \kappa A_z^2 && \text{if $v_z = 0$} \\
        J_z &= \frac{v_{z,\text{max}}^2}{2\kappa^2} && \text{if $z=0$} \text{.}
    \end{aligned}
\eeq

Now assume that the coordinates $(R, z, v_{\phi}, v_R, v_z)$ are offset by
$(\Delta R, \Delta z, \Delta v_{\phi}, \Delta v_R, \Delta v_z)$. Then, the
induced error in $J_{\phi}$ is, regardless of phase,
\beq\label{eq:induced_Jphi}
\frac{\Delta J_{\phi}}{J_{\phi}} = \frac{\Delta R}{R_g}\text{.}
\eeq
For $J_R$, the induced error is,
\beq\label{eq:induced_JR}
\begin{aligned}
    \frac{\Delta J_R}{J_R} &= \frac{2\Delta R}{A_R} + 
        \left(\frac{\Delta v_R}{v_{R,\text{max}}}\right)^2 && \text{if $v_R = 0$} \\
    \frac{\Delta J_R}{J_R} &= \frac{2\Delta v_R}{v_{R,\text{max}}} + 
        \left(\frac{\Delta R}{A_R}\right)^2  && \text{if $R=R_g$} \text{.}
\end{aligned}
\eeq
For $J_{z}$, the induced error is,
\beq\label{eq:induced_Jz}
\begin{aligned}
    \frac{\Delta J_z}{J_z} &= \frac{2\Delta z}{A_z} + 
        \left(\frac{\Delta v_z}{v_{z,\text{max}}}\right)^2 && \text{if $v_z = 0$} \\
    \frac{\Delta J_z}{J_z} &= \frac{2\Delta v_z}{v_{z,\text{max}}} + 
        \left(\frac{\Delta z}{A_z}\right)^2  && \text{if $z=0$} \text{.}
\end{aligned}
\eeq

Since most of the time stars will be at maximum amplitude (i.e. turnaround) in both $R$ and $z$, we can approximate the order of magnitude of the systematic error in the actions by
\beq\label{eq:Ji_err_mosttime}
\begin{split}
\frac{\Delta J_{\phi}}{J_{\phi}} &= \frac{\Delta R}{R_g} \\
\frac{\Delta J_{R}}{J_{R}} &= \frac{2\Delta R}{A_R} \\
\frac{\Delta J_{z}}{J_{z}} &= \frac{2\Delta z}{A_z} \text{,}
\end{split}
\eeq
where we have also ignored second order terms.

In the remainder of this section, we compare our analytic estimates of the effect on actions of a midplane offset against numerical calculations. A numerical evaluation of the effect of velocity offsets on actions is deferred to future work, as we discuss in Section~\ref{ssec:lsr_var}.
